---
title: "Simple Quarto Test"
format:
  html:
    toc: true
    toc-depth: 3
    code-link: true
    code-copy: true
    code-fold: false
---

# Links

General infos:
- https://arrow.apache.org/docs/r/

Code from: 
- https://rpubs.com/asgr/arrowdataset

# Installation and libs

```{r installationLibs}
if(FALSE){
  install.packages("arrow")
}

library(arrow)
library(data.table)
library(dplyr)
library(foreach)
library(microbenchmark)
```

# Test arrow
## Create test-data

```{r createTestData}
CSV_dir = paste0(tempdir(),'/CSV_DB/')
dir.create(CSV_dir)

set.seed(666)
dummy = foreach(i = 1:100) %do% {
  tempDT = data.table(let=letters[sample(26,1e3,TRUE)], num=rnorm(1e3), loc=i)
  fwrite(tempDT, paste0(CSV_dir,'examp_',formatC(i, width=3,flag=0),'.csv')) # files like examp_001.csv 
}
list.files(CSV_dir)
```

## Read data

```{r readData}
CSV_DB = arrow::open_dataset(CSV_dir, format='csv')

output1 = CSV_DB %>% #output table will be output1, input is the CSV_DB
  filter(let == 'a') %>% #filter all rows to where let =='a'
  collect %>% #collect actually creates the result
  setDT # setDT turns it into a data.table (which I prefer)
output2 = CSV_DB %>% filter(loc == 2) %>% collect %>% setDT #we put it on one line for compactness
output3 = CSV_DB %>% filter(num > 0) %>% collect %>% setDT

print(output1)
print(output2)
print(output3)

```

## Other operations on the data

```{r subsetting}
output4 = CSV_DB %>%
  filter(num > 0, let=='a') %>% #a more complex filter
  select(num, let) %>% # just num/let columns to be selected (in different order)
  collect %>% 
  setDT
print(output4)
```

```{r mutating}
output5 = CSV_DB %>%
  filter(num > 0, let=='a') %>% #a more complex filter
  mutate(mut = num*loc) %>% #pretty much any function you like here
  select(num, loc, mut) %>% #select some columns we care about
  collect %>%
  setDT

print(output5)
```
# Parquet

For speed (and bigger files) we might want to convert the database to a Parquet backend:

```{r writeParquet}
Parq_dir = paste0(tempdir(),'/Parq_DB/')

write_dataset(CSV_DB, Parq_dir, format='parquet', partitioning='loc') #partitioning tells it how to split the database
list.files(Parq_dir)

```

There are 100 folders with the binary files in it

```{r readParquetFiles}
Parq_DB = open_dataset(Parq_dir, format='parquet')
print(Parq_DB)

```

```{r readParquetData}
output1b = Parq_DB %>% #output table will be output1, input is the Parq_DB
  filter(let == 'a') %>%
  collect %>% 
  setDT
output2b = Parq_DB %>% filter(loc == 2) %>% collect %>% setDT
output3b = Parq_DB %>% filter(num > 0) %>% collect %>% setDT

print(output1b)
print(output2b)
print(output3b)
```

So, which is faster: csv or parquet?! Let’s see!

```{r benchmark1}
microbenchmark(
  {output3 = CSV_DB %>%filter(num > 0) %>% collect %>% setDT},
  {output3b = Parq_DB %>% filter(num > 0) %>% collect %>% setDT}
)

```

Where you will see big speedups is if you partition based on what you are likely to search on. Before we used `loc` new `let`. 

```{r repartitioning}
Parq_dir2 = paste0(tempdir(),'/Parq_DB2/')

write_dataset(CSV_DB, Parq_dir2, format='parquet', partitioning='let')
Parq_DB2 = open_dataset(Parq_dir2, format='parquet')
print(Parq_DB2) #should be 101 files, because there will be a parquet metadata file

output1c = Parq_DB2 %>% filter(let == 'a') %>% collect %>% setDT
output2c = Parq_DB2 %>% filter(loc == 2) %>% collect %>% setDT
output3c = Parq_DB2 %>% filter(num > 0) %>% collect %>% setDT
```

```{r benchmark2}
microbenchmark(
  {output1 = CSV_DB %>% filter(let == 'a') %>% collect %>% setDT},
  {output1b = Parq_DB %>% filter(let == 'a') %>% collect %>% setDT},
  {output1c = Parq_DB2 %>% filter(let == 'a') %>% collect %>% setDT}
)
```

# Convert an In-Memory Table to Arrow Dataset

f you already have your target arrow Dataset loaded into an R object then you can directly write this out to an arrow Dataset. This might be the case where you have a machine with a lot of memory, but you want to share the Dataset so people using it with a small memory laptop over an internet connection can still do useful things.

```{r inMemoryToParquet}
tempDT = data.table(let=letters[sample(26,1e5,TRUE)], num=rnorm(1e5), loc=rep(1:100,each=1e3))
Parq_dir3 = paste0(tempdir(),'/Parq_DB3/')

write_dataset(tempDT, Parq_dir3, format='parquet', partitioning=c('let','loc'))
Parq_DB3 = open_dataset(Parq_dir3, format='parquet')
print(Parq_DB3) #should be 101 files, because there will be a parquet metadata file
```

